Comparing the Direction and Degree of Selection on Mutations Between SARS-CoV-2 Phylogenetic Clades

### 2022-10-25 notebooks/expected_and_actual_counts_ratio.Rmd
Reorder code to characterize and filter data

---
title: "Comparing the Direction and Degree of Selection on Mutations Between SARS-CoV-2 Phylogenetic Clades"
output: html_document
date: "2022-10-17"
---

@@

## SARS-CoV-2 Mutation Data

@@

Next, take a look at the first 5 rows of the data.

```{r head_data}
head(SARS_mutation, n=5)
```

```{r describe_data}
# How many rows? How many columns?
dim(SARS_mutation)

# How many phylogenetic clades?
length(unique(SARS_mutation$clade))

# How many rows for each subset? Data from all countries, only England or only the United States.
data.frame(all=nrow(SARS_mutation[which(SARS_mutation$subset=="all"),]), England=nrow(SARS_mutation[which(SARS_mutation$subset=="England"),]), USA=nrow(SARS_mutation[which(SARS_mutation$subset=="USA"),]))

# How many nucleotide sites?
length(unique(SARS_mutation$nt_site))

# How many mutations to exclude for each subset?
data.frame(all=nrow(SARS_mutation[which(SARS_mutation$subset=="all" & SARS_mutation$exclude=="TRUE"),]), England=nrow(SARS_mutation[which(SARS_mutation$subset=="England" & SARS_mutation$exclude=="TRUE"),]), USA=nrow(SARS_mutation[which(SARS_mutation$subset=="USA" & SARS_mutation$exclude=="TRUE"),]))
```

## Filter and Prepare the Data

@@

To compare mutations between phylogenetic clades, this notebook will plot volcano plots with p-value on the Y-axis and fold-change on the X-axis. Prior to filtering the data for contingency table analyses, calculate the ratio of observed to expected counts and add a pseudocount to prevent infinities when fold-change is calculated.

```{r pseudocount}
# Calculate the ratio of actual counts (i.e., observed) to expected counts (i.e., expected) with pseudocount of 0.5.
filtered_SARS_mutation$observed_expected <- (filtered_SARS_mutation$actual_count + 0.5)/(filtered_SARS_mutation$expected_count + 0.5)
```

## Visualize Filtered Data

@@

```{r subset_by_clade}
# Create a subset of the data using only the clades that are going to be compared.
clade_SARS_mutation <- filtered_SARS_mutation[which(filtered_SARS_mutation$clade=="20A" | filtered_SARS_mutation$clade=="20B"),]

# Identify unique clades in the dataset.
clades <- unique(clade_SARS_mutation$clade)
```

>>>>

---
title: "Comparing the Direction and Degree of Selection on Mutations Between SARS-CoV-2 Phylogenetic Clades"
output: html_document
date: Sys.time()
---

@@

@@

## Define Filtering and Analysis Parameters

```{r parameters}
# The subset column describes what countries the sequence data came from. Identify what the subset options are.
unique(SARS_mutation$subset)

# Define what subset of the data to analyze.
country <- "all"

# Print the clades that are in the data set (i.e., what clades are possible to include in comparisons).
unique(SARS_mutation$clade)

# Create a list of the clades to include in comparisons.
clades <- c("20A","20B")

# Define a minimum threshold for the number of expected counts. All mutations below this threshold will be removed.
min_expected <- 10

# To avoid infinities when calculating fold-change, add a pseudocount to the observed and expected counts.
pseudo <- 0.5
```

## Filter and Prepare the SARS-CoV-2 Mutation Data

@@

To compare mutations between phylogenetic clades, this notebook will plot volcano plots with p-value on the Y-axis and fold-change on the X-axis. Prior to filtering the data for contingency table analyses, calculate the ratio of observed to expected counts and add a pseudocount to prevent infinities when fold-change is calculated. Mutations with a small count (i.e., 1) are going to be affected by the pseudocount more than mutations with a large count (i.e., 100).

```{r pseudocount}
# Calculate the ratio of actual counts (i.e., observed) to expected counts (i.e., expected) with pseudocount of 0.5.
filtered_SARS_mutation$observed_expected <- (filtered_SARS_mutation$actual_count + pseudo)/(filtered_SARS_mutation$expected_count + pseudo)
```

## Characterize the Filtered Data

Take a look at the first 5 rows of the data.

```{r head_data}
head(filtered_SARS_mutation, n=5)
```

```{r describe_data}
# How many rows? How many columns?
dim(filtered_SARS_mutation)

# How many phylogenetic clades?
length(unique(filtered_SARS_mutation$clade))

# How many nucleotide sites?
length(unique(filtered_SARS_mutation$nt_site))
```

## Visualize the Filtered Data

@@

```{r subset_by_clade}
# Create a subset of the data using only the clades that are going to be compared.
clade_SARS_mutation <- filtered_SARS_mutation[filtered_SARS_mutation$clade %in% clades,]
```
